---
title: "Lab 02 - Global plastic waste"
author: "Kaden Zhu"
output: 
  tufte::tufte_html:
    css: ../lab.css
    tufte_variant: "envisioned"
    highlight: pygments
  tufte::tufte_handout:
    latex_engine: xelatex
    highlight: pygments
    keep_tex: true
    citation_package: natbib
link-citations: true
---
```{r include = FALSE}
knitr::opts_chunk$set(eval=TRUE)
```

## Introduction

Plastic pollution is a major and growing problem, negatively affecting oceans and wildlife health. [Our World in Data](https://ourworldindata.org/plastic-pollution) has a lot of great data at various levels including globally, per country, and over time. For this lab we focus on data from 2010.

Additionally, National Geographic ran a data visualization communication contest on plastic waste as seen [here](https://www.nationalgeographic.org/funding-opportunities/innovation-challenges/plastic/dataviz/).

This lab builds on skills from Lab 01. You'll get more practice with ggplot2, learn about different types of visualizations, and explore a real-world dataset.

**Estimated time:** 90-120 minutes

## Learning Goals

By the end of this lab, you will be able to:

- Create and interpret histograms and density plots
- Use faceting to compare distributions across groups
- Adjust transparency (alpha) to improve visualizations
- Create and interpret box plots and violin plots
- Make scatterplots to visualize relationships between variables
- Modify aesthetics (color, fill) to enhance plots
- Build complex multi-layered visualizations

## Prerequisites

Before you begin, make sure you have:

- Completed Lab 01
- Watched the data visualization lecture
- Forked the lab-instructions repository
- Feel comfortable with the knit/commit/push workflow

# Getting Started

Navigate to your forked `lab-instructions` repository in JupyterHub, open the `Lab02` folder, and open the R Markdown document `lab-02.Rmd`.

## Verify Your Setup

Before proceeding:

1. Open `lab-02.Rmd` in RStudio
2. Check that the Git pane shows YOUR username (not the course organization)
3. Click **Knit** to make sure the document compiles

**If you see errors or can't find the file, ask for help before continuing!**

## Packages

We'll use the **tidyverse** package for this analysis. Run the following code in the Console to load this package.
```{r load-packages, message=FALSE, eval=TRUE}
library(tidyverse)
```

## Data

The dataset for this assignment can be found as a csv file in the `data` folder of your repository. You can read it in using the following.
```{r load-data, message=FALSE, eval=TRUE}
plastic_waste <- read_csv("data/plastic-waste.csv")
```

The variable descriptions are as follows:

-   `code`: 3 Letter country code
-   `entity`: Country name
-   `continent`: Continent name
-   `year`: Year
-   `gdp_per_cap`: GDP per capita constant 2011 international \$, rate
-   `plastic_waste_per_cap`: Amount of plastic waste per capita in kg/day
-   `mismanaged_plastic_waste_per_cap`: Amount of mismanaged plastic waste per capita in kg/day
-   `mismanaged_plastic_waste`: Tonnes of mismanaged plastic waste
-   `coastal_pop`: Number of individuals living on/near coast
-   `total_pop`: Total population according to Gapminder

# Warm Up

Before diving into the exercises, let's make sure you're familiar with the RStudio environment and the data.

**Question 1:** Without looking, can you name the four panes in RStudio and briefly describe their purpose?

**Your answer:** 

There is the Source Pane where you can edit the source document such as R Markdown files or any scripts you can also knit this and view it in pdf or html. Then there is the console and terminal pane where you can load packages or enter other commands and see the output in the console/terminal. There is also thehome pane with your packages, help which includes R resources manuals docs for Rstudio and more reference material, you also have viewer which you can attach a file to read, and plots that you have which you can display. Lastly you have the environment and history pane, we use this mostly for viewing the data we've loaded, our history, and manage github versioning with viewing the diff commiting and pushing. 

**Question 2:** Verify that the dataset has loaded into the Environment pane. How many observations are in the dataset?

**Hint:** Look in the Environment pane, or run `nrow(plastic_waste)` in the Console, or click on the dataset name to view it.

**Your answer:** The dataset has 240 observations.

**Question 3:** Have a quick look at the data using `View(plastic_waste)` in the Console. You'll notice that some cells contain `NA`. What does `NA` mean in R?

**Hint:** Run `?NA` in the Console to see the documentation.

**Your answer:**

In R, NA stands for “Not Available” and represents a missing value in the data. It is a special indicator used when a value is unknown, undefined, or not recorded. If you perform any numerical computations involving NA, the result will typically also be NA unless you explicitly handle missing values. R includes different typed versions of NA, such as NA_integer_, NA_real_, NA_complex_, and NA_character_, which correspond to different data types. The function anyNA() can be used to check for missing values in objects like atomic vectors, lists, pairlists, and even NULL.

**✓ Checkpoint:** Make sure the data loaded correctly and you can see it in your Environment pane before continuing.

🧶 ✅ ⬆️ **Knit, commit with message "Completed warm up", and push your changes.**

---

# Exercises

## Exploring Distributions

Let's start by taking a look at the distribution of plastic waste per capita in 2010.
```{r plastic_waste_per_cap-hist, eval=TRUE}
ggplot(data = plastic_waste, aes(x = plastic_waste_per_cap)) +
  geom_histogram(binwidth = 0.2)
```

One country stands out as an unusual observation at the top of the distribution. One way of identifying this country is to filter the data for countries where plastic waste per capita is greater than 3.5 kg/person.
```{r plastic_waste_per_cap-max, eval=TRUE}
plastic_waste %>%
  filter(plastic_waste_per_cap > 3.5)
```

Did you expect this result? Trinidad and Tobago has unusually high plastic waste per capita. You might research why this is the case!

---

## Exercise 1

Plot, using histograms, the distribution of plastic waste per capita faceted by continent. What can you say about how the continents compare to each other in terms of their plastic waste per capita?

**Reminder:** Faceting creates separate plots for each group. Use `facet_wrap(~variable)` to facet by a categorical variable.
```{r plastic-waste-continent, eval=TRUE}
# Remember to change eval=TRUE to eval=TRUE after filling in the blanks!
ggplot(data = plastic_waste, 
       aes(x = plastic_waste_per_cap)) +
  geom_histogram(binwidth = 0.2) +
  facet_wrap(~continent)
```

**Interpretation:** Compare the continents in terms of their plastic waste per capita. Consider:

- Which continent has the highest typical plastic waste per capita?
- Which has the lowest?
- Which continent shows the most variation?

**Your answer:**

The continent that has the highest typical plastic waste per cap is North America, yes asia and europe have similar bars on the lower end of plastic waste per capita, but you see clearly NA extends to the right most out of all the other countries above 3.5

Africa has lowest typical plastic waste per capita, although South americas bars are shorter they are more to the right and have more spread, Africa's highest and most typical is at the left most plastic waste per cap.

North America also shows the most variation. Most of its countries fall between 0 and 0.7, but there are a few much higher values above 3, creating a wide spread in the distribution.

---

## Density Plots

Another way of visualizing numerical data is using density plots. Density plots show the distribution as a smooth curve rather than bars.
```{r plastic_waste_per_cap-dens, eval=TRUE}
ggplot(data = plastic_waste, aes(x = plastic_waste_per_cap)) +
  geom_density()
```

We can compare distributions across continents by coloring density curves by continent.
```{r plastic_waste_per_cap-dens-color, eval=TRUE}
ggplot(data = plastic_waste, 
       mapping = aes(x = plastic_waste_per_cap, 
                     color = continent)) +
  geom_density()
```

The resulting plot may be a little difficult to read, so let's also fill the curves in with colors as well.
```{r plastic_waste_per_cap-dens-color-fill, eval=TRUE}
ggplot(data = plastic_waste, 
       mapping = aes(x = plastic_waste_per_cap, 
                     color = continent, 
                     fill = continent)) +
  geom_density()
```

The overlapping colors make it difficult to tell what's happening with the distributions in continents plotted first. We can change the transparency level of the fill color to help with this. The `alpha` argument takes values between 0 and 1: 0 is completely transparent and 1 is completely opaque.
```{r plastic_waste_per_cap-dens-color-fill-alpha, eval=TRUE}
ggplot(data = plastic_waste, 
       mapping = aes(x = plastic_waste_per_cap, 
                     color = continent, 
                     fill = continent)) +
  geom_density(alpha = 0.7)
```

This still doesn't look great...

---

## Exercise 2

Recreate the density plots above using a different (lower) alpha level that works better for displaying the density curves for all continents.

**Try alpha values like 0.3, 0.4, or 0.5 to see which works best.**
```{r plastic-waste-density-alpha, eval=TRUE}
# Remember to change eval=TRUE to eval=TRUE!
ggplot(data = plastic_waste, 
       mapping = aes(x = plastic_waste_per_cap, 
                     color = continent, 
                     fill = continent)) +
  geom_density(alpha = 0.25)
```

**Which alpha level did you choose and why?**
I chose 0.25 because you can see the overlapping lines that were covered behind south america clearer but you can also distinguish them apart because the alpha isn't too bare that everything will be so light and indistinguishable. 

---

## Exercise 3

Describe why we defined the `color` and `fill` of the curves by mapping aesthetics of the plot (inside `aes()`) but we defined the `alpha` level as a characteristic of the plotting geom (inside `geom_density()`).

**Hint:** Think about what varies by data vs. what is constant across all data points.

**Your answer:**
We mapped color and fill inside aes() because they vary based on a data variable, such as continent. This allows each group in the data to have a different color and fill automatically.
Alpha was defined outside of aes() because it is not meant to vary by data. It is a constant setting that applies equally to all curves in the plot. If we placed alpha inside aes(), it would try to map transparency to a variable, which is not what we want.


🧶 ✅ ⬆️ **Knit, commit with message "Completed Exercises 1-3", and push your changes.**

---

## Box Plots and Violin Plots

Another way to visualize this relationship is using side-by-side box plots.
```{r plastic_waste_per_cap-box, eval=TRUE}
ggplot(data = plastic_waste, 
       mapping = aes(x = continent, 
                     y = plastic_waste_per_cap)) +
  geom_boxplot()
```

---

## Exercise 4

Convert your side-by-side box plots from the previous task to [violin plots](http://ggplot2.tidyverse.org/reference/geom_violin.html). What do the violin plots reveal that box plots do not? What features are apparent in the box plots but not in the violin plots?

**Hint:** Simply replace `geom_boxplot()` with `geom_violin()`.
```{r plastic-waste-violin, eval=TRUE}
# Remember to change eval=TRUE to eval=TRUE!
ggplot(data = plastic_waste, 
       mapping = aes(x = continent, 
                     y = plastic_waste_per_cap)) +
  geom_violin()
```

**What do violin plots reveal that box plots do not?**
Violin plots reveal the full distribution and density of the data and visualizes it with its wideness. Box plots show the median, interquartile range, and potential outliers, they do not show how the data are distributed within that range. A violin plot allows us to see where values are more concentrated by showing the density at different levels. This helps us understand how many countries cluster around certain plastic waste per capita values, rather than only seeing summary statistics.

**What features are apparent in box plots but not in violin plots?**

Box plots clearly show the median as a line inside the box, the interquartile range, and outliers. In a standard violin plot, the exact median and quartiles are not specifically marked unless they are added separately.Violin plots show density, they don't make summary statistics right in the visualization like the median as and not immediately clear as box plots do.
---

## Scatterplots

Now let's explore relationships between variables using scatterplots.

## Exercise 5

Visualize the relationship between plastic waste per capita and mismanaged plastic waste per capita using a scatterplot. Describe the relationship.

**Reminder:** Use `geom_point()` to create scatterplots.
```{r plastic-waste-mismanaged, eval=TRUE}
# Remember to change eval=TRUE to eval=TRUE!
ggplot(data = plastic_waste, 
       mapping = aes(x = plastic_waste_per_cap, 
                     y = mismanaged_plastic_waste_per_cap)) +
  geom_point()
```

**Describe the relationship between plastic waste per capita and mismanaged plastic waste per capita:**

There is a postitive relationship between plastic waste per capita and mismanaged plastic waste per capita most countries cluster in the lower end of plastic waste per capita meaning they have both low plastic waste per capita and low mismanaged waste. There's a few higher values where countries with greater plastic consumption also show higher levels of mismanaged waste.
---

## Exercise 6

Color the points in the scatterplot by continent. Does there seem to be any clear distinctions between continents with respect to how plastic waste per capita and mismanaged plastic waste per capita are associated?

**Hint:** Add `color = continent` to your `aes()` mapping.
```{r plastic-waste-mismanaged-continent, eval=TRUE}
# Remember to change eval=TRUE to eval=TRUE!
ggplot(data = plastic_waste, 
       mapping = aes(x = plastic_waste_per_cap, 
                     y = mismanaged_plastic_waste_per_cap,
                     color = continent)) +
  geom_point()
```

**Are there clear distinctions between continents?**
Most countries overlap heavily on the lower range of plastic waste per cap and they all cluster near the low values america stand out just slightly because it has an extremem outlier but China seems to have higher mismanaged plastic waste per capita compared to the others and the highest point on the lower end. and Europe has a little but lower mismanaged plastic waste per cap. 


🧶 ✅ ⬆️ **Knit, commit with message "Completed Exercises 4-6", and push your changes.**

---

## Exercise 7

Visualize the relationship between plastic waste per capita and total population as well as plastic waste per capita and coastal population. You will need to make two separate plots. Do either of these pairs of variables appear to be more strongly linearly associated?

**Plot 1: Plastic waste per capita vs. total population**
```{r plastic-waste-population, eval=TRUE}
# Remember to change eval=TRUE to eval=TRUE!
ggplot(data = plastic_waste, 
       mapping = aes(x = plastic_waste_per_cap, 
                     y = total_pop)) +
  geom_point()
```

**Plot 2: Plastic waste per capita vs. coastal population**
```{r plastic-waste-coastal, eval=TRUE}
# Remember to change eval=TRUE to eval=TRUE!
ggplot(data = plastic_waste, 
       mapping = aes(x = plastic_waste_per_cap, 
                     y = coastal_pop)) +
  geom_point()
```

**Which pair of variables appears to be more strongly linearly associated?**

Between the two pairs, plastic waste per capita and coastal population appear slightly more linearly associated than plastic waste per capita and total population neither has a very strong positive or negative relationship they both show heavy clustering at low values

🧶 ✅ ⬆️ **Knit, commit with message "Completed Exercise 7", and push your changes.**

---

## Challenge Exercise

## Exercise 8

Recreate the following plot, and interpret what you see in context of the data.
```{r echo=FALSE, message=FALSE, eval=TRUE, warning=FALSE}
plastic_waste %>% 
  mutate(coastal_pop_prop = coastal_pop / total_pop) %>%
  filter(plastic_waste_per_cap < 3) %>%
  ggplot(aes(x = coastal_pop_prop, y = plastic_waste_per_cap, color = continent)) + 
    geom_point() +
    geom_smooth(color = "black") +
    scale_color_viridis_d() +
    labs(x = "Coastal population proportion (Coastal / total population)", 
         y = "Plastic waste per capita ", 
         color = "Continent",
         title = "Plastic waste vs. coastal population proportion",
         subtitle = "by continent") +
    theme_minimal()
```

**This plot requires several steps:**

1. Create a new variable `coastal_pop_prop` = `coastal_pop` / `total_pop`
2. Filter for observations where `plastic_waste_per_cap` < 3
3. Create a scatterplot with the new variable on x-axis
4. Add a smooth curve with `geom_smooth()`
5. Use `scale_color_viridis_d()` for the color scale
6. Add appropriate labels

**Build it step by step:**
```{r recreate-plot, eval=TRUE}
# Remember to change eval=TRUE to eval=TRUE!
plastic_waste %>% 
  mutate(coastal_pop_prop = coastal_pop / total_pop) %>%
  filter(plastic_waste_per_cap < 3) %>%
  ggplot(aes(x = coastal_pop_prop, y = plastic_waste_per_cap, color = continent)) + 
    geom_point() +
    geom_smooth(color = "black") +
    scale_color_viridis_d() +
    labs(x = "Coastal population proportion coastal/total pop", 
         y = "Plastic Waste Per Cap", 
         color = "Continent",
         title = "Plastic Waste vs Coastal population proportion",
         subtitle = "by Continent") +
    theme_minimal()
```

**Interpretation:** What does this plot reveal about the relationship between coastal population proportion and plastic waste per capita? Does the relationship differ across continents?

theres a slight positive relationship between coastal population proportion and plastic waste per capita, as the proportion of people living near coastal areas increases the plastic waste per capita tends to increase, the trend line looks like a smooth increase so the higher the costal pop out of total pop the higher plastic waste per cap


🧶 ✅ ⬆️ **Knit, commit with message "Completed Exercise 8", and push your changes.**

---

## Common Errors and Troubleshooting

**Visualization errors:**

- **Plot not showing** → Make sure you changed `eval=TRUE` to `eval=TRUE` in the chunk options
- **Error: "object not found"** → Check spelling of variable names (case-sensitive!)
- **Error: "could not find function 'geom_violin'"** → Make sure tidyverse is loaded
- **Weird looking plots** → Try different alpha values, binwidths, or figure sizes

**Data errors:**

- **Can't load data** → Make sure the `data` folder is in the same directory as your .Rmd file
- **NA values causing issues** → This is normal for this dataset; R handles them appropriately in plots

**Git/GitHub errors:**

- **Can't push** → Make sure you committed first
- **Changes not showing on GitHub** → Check that you pushed to YOUR fork, not the course repo


**To submit to Canvas:**

1.  In RStudio, click the **Knit** dropdown menu (next to the Knit button)
2.  Select **Knit to tufte_handout** to generate a PDF
3.  Download the PDF file from the Files pane
4.  Upload the PDF to Canvas

**Great work! Make sure all your changes are committed and pushed to GitHub.**